Compositional Quantitative Resource Analysis of C Programs

### 2014-06-15 paper/main.tex
work on section 2

\section{Amortized Analysis for C Programs}
\label{sec:inform}

\subsection{Amortized Analysis}
The idea that drives the design of our system is amortized analysis
[XXX].  Say that a program $p$ evaluates on a program state $H$ by
consuming $n$ resource units, we note that $(p, H) \Downarrow_n$.  The
basic idea of amortized analysis is to define a potential function
$\Phi$ that maps program states to non-negative numbers, such that
$\Phi(H) \ge n$.

Since imperative programs can be composed we must enrich
the judgement $(\_, \_) \Downarrow_{\_}$ to talk about the heap
resulting from the program execution $H'$.  So we now write
$(p, H) \Downarrow_n H'$ to say that $H'$ results from $p$'s
execution.  We now use two potential functions, one
that applies before the execution, and one that applies
after.  The two functions must respect the natural relation
$\Phi(H) \ge n + \Phi'(H')$.  Intuitively speaking, $\Phi$ must
give enough \emph{potential} to, both, finish the computation
and pay for $\Phi'$ on the resulting heap. That way, if $(H, p_1) \Downarrow_n H'$
and $(H', p_2) \Downarrow_m H''$, we get $\Phi(H) \ge
n + \Phi'(H')$ and $\Phi'(H') \ge m + \Phi''(H'')$, which
can be composed to give $\Phi(H) \ge (n + m) + \Phi''(H'')$.
What we have observed is that, if we define $\Phi \vdash p \dashv \Phi'$
to be
$$
\forall H\,n\,H',~(H, p)\Downarrow_n H' \implies \Phi(H) \ge
n + \Phi'(H'),
$$
then we get the following familiar looking rule:
$$
\RuleNolabel
{\Phi \vdash p_1 \dashv \Phi' \\ \Phi' \vdash p_2 \dashv \Phi''}
{\Phi \vdash p_1; p_2 \dashv \Phi''}
.
$$
Similarly, other language constructs contrain the potential
functions to respect more rules that look very similar to
Hoare logic or effect systems rules.  By fixing the shape of
a potential function $\Phi$, it becomes possible to leverage
existing tools to help finding a pair of potential functions
for an arbitrary program.  This is explained in further details
in the following sections.

We say that the analysis is amortized because the use of a
potential function $\Phi$ gives the flexibility to amortize some
operations.  For example, if heaps are simple mappings from
$\{\code x, \code y\}$ to natural numbers, we have the following
judgement (provided that $H~\code x \ge H~\code y \ge 0$ initially)
$$
(\Phi := \lambda H.~2\cdot(H~\code x) + 10)
\vdash \code x = \code x - \code y \dashv
(\Phi' := \lambda H'.~2\cdot(H'~\code x) + 2\cdot(H'~\code y) + 10).
$$
This must be understood as a motion of potential from $\code x$
to $\code y$. If some resource is consumed $\code y$ times afterwards, we
can now use $\code y$'s potential to pay for it. But interestingly, the
initial potential function $\Phi$ only talks about~$\code x$: the potential
function allowed us to impart the cost of the subsequent resource
consumptions on the variable~$\code x$.
This flexibility combined with intelligent decisions on potential
motions allow our tool to discover linear invariants for surprisingly
convoluted programs.

@@

\section{Catalog of Automatically Analyzed Programs}
\lable{app:cat}

@@

>>>>

\section{Informal Account}
\label{sec:inform}

In this section, we informally introduce the quantiatative program
logic and the automatic amortized analysis for for Clight programs.
The purpose of this section is to build some intuition that should
make the following sections, which contain more details, easy to read.

\subsection{Quantitative Hoare Logic}

The idea that drives the design of our system is amortized analysis
[XXX].  Assume that a program $p$ executes on a stating state
$\state_0$ and consumes $n$ resource units of some quantity we are
interested in.  We denote that by writing $(p, \state_0)
\Downarrow_n$.  The basic idea of amortized analysis is to define a
potential function $\Phi$ that maps program states to non-negative
numbers and to show that $\Phi(\state) \geq n$ if $\state$ is a
program state such that such that $(p, \state) \Downarrow_n$.

Since programs can be composed we must enrich the evaluation judgement
$(p, \state) \Downarrow_{n}$ to talk about the state resulting from
the program execution.  We write $(p, \state) \Downarrow_n \state'$ to
say that $\state'$ is the program state after the execution of $p$ in
starting state $\sigma$.  We now use two potential functions, one that
applies before the execution, and one that applies after.  The two
functions must respect the relation $\Phi(\state) \ge n +
\Phi'(\state')$ for all states $\state$ and $\state'$ such that $(p,
\state) \Downarrow_n \state'$.  Intuitively, $\Phi(\state)$ must
provide enough \emph{potential} for, both, paying for resource cost of
the computation and paying for the potential $\Phi'(\state')$ on the
resulting state $\state'$. That way, if $(\state, p_1) \Downarrow_n
\state'$ and $(\state', p_2) \Downarrow_m \state''$, we get
$\Phi(\state) \ge n + \Phi'(\state')$ and $\Phi'(\state') \ge m +
\Phi''(\state'')$.  This can be composed to $\Phi(\state) \ge (n + m)
+ \Phi''(\state'')$.  Note that the initial potential function $\Phi$
provides an upper bound on the resource consumption of the whole
program.  What we have observed is that, if we define $\htriple{\Phi}
{p}{\Phi'}$ to mean
$$
\forall \state, n, \state' : (\state, p) \Downarrow_n \state' \implies \Phi(\state) \ge
n + \Phi'(\state')
$$
then we get the following familiar looking rule.
$$
\RuleNolabel
{\htriple{\Phi}{p_1}{\Phi'} \\ \htriple{\Phi'}{p_2}{\Phi''}}
{\htriple{\Phi} {p_1; p_2}  {\Phi''}}
$$
%
Similarly, other language constructs lead to rules for the potential
functions that look very similar to Hoare logic or effect systems
rules.  This rules enable to interactively reason about resource usage
in a flexible and compositional way, which, as a side effect, produces
a certificate for the derived resource bound.

\paragraph{Automatic Amortized Analysis}

By fixing the shape of a potential function $\Phi$, it becomes
possible to leverage existing tools to help finding a pair of
potential functions for an arbitrary program.  This is explained in
further details in the following sections.

We say that the analysis is amortized because the use of a potential
function $\Phi$ gives the flexibility to amortize some operations.
For example, if program states are simple mappings from variables to
natural numbers, we have the following judgement (provided that
initially $\state(\code x) \ge \state(\code y) \ge 0$).
$$
\htriple
{\lambda \state . \, 2 {\cdot} \state (\code x) + 10}
{\code{ x = x -  y}}
{\lambda \state . \, 2 {\cdot} \state(\code x) + 2 {\cdot} \state(\code y) + 10}
$$
This must be understood as a motion of potential from the
variable~$\code x$ to the variable~$\code y$. If some resource is
consumed $\code y$ times afterwards, we can then use $\code y$'s
potential to pay for it. However, the initial potential function
$\Phi$ only refers to the variable~$\code x$: the potential function
allows us to pass on the cost of the subsequent resource consumption
to~$\code x$.  This flexibility combined with intelligent decisions on
potential motions allow our tool to discover linear invariants for
surprisingly convoluted programs.

@@

\section{Catalog of Automatically Analyzed Programs}
\label{app:cat}

@@

%%% Local Variables: 
%%% mode: latex
%%% mode: flyspell
%%% TeX-master: t
%%% End:
